---
title: "Diamonds analysis"
author:
  - "Oriade Simpson (s172084)"
  - "Pietro Lombardo (s231756)"
date: '2023-09-05'
output:
  pdf_document:
    keep_tex: true
    extra_dependencies: "subfig"
    includes:
      in_header: columns.tex
  html_document:
    df_print: paged
    css: columns.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

library(MASS)
library(ggplot2)
library(data.table)
library(mltools)
library(reshape)
library(reshape2)
library(rmarkdown)
library(tidyverse)
library(knitr)

data(diamonds)
# ctrl + alt + I
```


Contribution Table 
===
------------------------------------------------------------

Task          |     Oriade    |     Pietro    |
--------------|:-------------:|:-------------:|
**Student ID**|   s172084     |    s231756    |
------------- |-------------- | ------------- | 
Question 1    |       x       |               |   
------------- | ------------- | ------------- |
Question 2    |       x       |               | 
------------- | ------------- | ------------- |
Question 3    |               |       x       |  
------------- | ------------- | ------------- |
Question 4    |               |       x       |
------------- | ------------- | ------------- |
Exam Prob 1   |               |       x       | 
------------- | ------------- | ------------- |
Exam Prob 2   |       x       |               |
------------- | ------------- | ------------- |
Exam Prob 3   |       x       |               |
------------- | ------------- | ------------- |
Exam Prob 4   |               |       x       |
------------- | ------------- | ------------- |
Exam Prob 5   |               |       x       |
------------- | ------------- | ------------- |
Exam Prob 6   |       x       |               |

# 1) Description of the dataset

- Explain what your data is about. I.e. what is the overall problem of
interest?
- Provide a reference to where you obtained the data.
- Summarize previous analysis of the data. (i.e. go through one or two
of the original source papers and read what they did to the data and
summarize their results).
- You will be asked to apply (1) classification and (2) regression on your
data in the next report. For now, we want you to consider how this should
be done. Therefore:
Explain, in the context of your problem of interest, what you hope to
accomplish/learn from the data using these techniques?.
Explain which attribute you wish to predict in the regression based on
which other attributes? Which class label will you predict based on which
other attributes in the classification task?
If you need to transform the data in order to carry out these tasks, explain
roughly how you plan to do this.

One of these tasks (1)–(5) is likely more relevant than the rest and will be
denoted the main machine learning aim in the following. The purpose of the 
following questions, which asks you to describe/visualize the data, is to
allow you to reflect on the feasibility of this task.

# 2) Explanation of the attributes of the data

- Describe if the attributes are discrete/continuous, Nominal/Ordinal/Interval/
Ratio,
- Give an account of whether there are data issues (i.e. missing values or
corrupted data) and describe them if so.
- Include basic summary statistics of the attributes.

If your data set contains many similar attributes, you may restrict yourself to
describing a few representative features (apply common sense).

# 3) Data visualization

## Principal component analysis

The aim of the Principal Component Analysis ("PCA" from now on), is to simplify
the problem dimension by reducing the number of variables which explains the
behavior of the diamonds' price.

The PCA requires the standardization of the attributes so that the variability of
each of them is comparable. But such operation cannot be carried out for 
non-numeric values, like the three ordinal variables *cut*, *color*
and *clarity*. For these reason they are converted into ordinal numbers 
(from 1 to the upper level number) according to their ranking.

After these operations, the dataset appears as below:

```{r}
Dia <- as.data.frame(diamonds)
Dia$cut <- as.numeric(Dia$cut)
Dia$color <- as.numeric(Dia$color)
Dia$clarity <- as.numeric(Dia$clarity)

N <- dim(Dia)[1] # number of rows of the new one-hot-encoded dataframe
M <- dim(Dia)[2] # number of columns of the new one-hot-encoded dataframe
attributeNames <- colnames(Dia)

kable(head(Dia), caption = "Head of the diamonds dataset arranged for the PCA")
```

Now the dataset contains all numeric variables, so the next step is to compare
the standard deviations of each variable and check whether they are different.

Below the standard deviations of the variables are shown:

```{r}
stds <- apply(Dia, 2, sd)
print(round(stds,digits=2))
```

Since standard deviation of *price* is some orders of magnitude larger than the 
others, the dataset has to be standardized by subtracting the mean and dividing
by the standard deviation of the whole set of observations of that variable.
```{r}
D_pca <- t(apply(Dia, 1, "-", colMeans(Dia)))
D_pca <- t(apply(D_pca, 1, "*", 1 / stds))
D_pca <- as.data.frame(D_pca)
```

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.55\textwidth}"}
```{r, echo=FALSE, fig.width=5, fig.height=4}
S <- svd(D_pca)
diagS <- S$d
rho <- diagS^2 / sum(diagS^2)

threshold <- 0.9

xlimits <- c(1, M)
plot(rho,
     type = "o",
     main = "Variance explained by principal components",
     xlab = "Principal components",
     ylab = "Variance explained",
     xlim = xlimits,
     ylim = c(0, 1),
     col = "blue"
)

lines(cumsum(rho), type = "o", col = "orange")
lines(xlimits, c(threshold, threshold), lty = "dashed")

legend("right", # Define position
       legend = c("Cumulative", "Individual", "Threshold"), # Set strings for legend
       col = c("orange", "blue", "black"), lty = c(1, 1, 2), # Match appereance of lines
       cex = 1, bg = "lightblue"
) # Setup how the box looks (cex controls size)
```
:::

::: {.col data-latex="{0.05\textwidth}"}
\ 
:::

::: {.col data-latex="{0.40\textwidth}"}

The new standardized dataset can be now used to perform the Singular Value
Decomposition (SVD), which gives rise to three different matrices: $U$, $\Sigma$ 
and $V^T$. By the extraction of the diagonal of the matrix $\Sigma$,
it can be seen how much variance is explained by each principal component.
The cumulative explained variance should reach the percentage of 90% in order
to describe properly the main features of the dataset. 

The figure on the left shows that the first 4 principal components explain 
little less than the 90% of variance
:::
::::::

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.55\textwidth}"}
```{r, echo=FALSE, fig.width=5, fig.height=4}
Z <- S$u %*% diag(S$d)
V <- S$v

pcs <- 1:4
test <- as.data.frame(melt(data.table(V[, pcs])))
ggplot(test, aes(x = rep(1:10, length(pcs)), y = value, fill=variable)) +
  ggtitle("Principal directions interpreted in terms of features") +
  geom_bar(position="dodge", stat = "identity") +
  labs(fill="PC", x = "Attributes", y = "Component coefficients")
```
:::

::: {.col data-latex="{0.05\textwidth}"}
\ 
:::

::: {.col data-latex="{0.40\textwidth}"}
The matrix $V^T$ contains the ten deca-dimensional vectors defining the principal 
components (PCs). Focusing on the first four PCs, they contains the weights
associated to each of original component. The figure on the left shows
the valuea of the weights. It can be noticed that the first PC mainly describes
the dimensional quantities of the diamonds (*carat*, size in *x*, *y*, *z*) and
the *price*. It seems that these five characteristics alone explain half of the
variability of the diamonds. As for the second PC, it focuses more on the
quality characteristics (*cut*, *color*, *clarity*) and on the *table*.
:::
::::::

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.55\textwidth}"}
```{r, echo=FALSE, fig.width=5, fig.height=5}
i <- 1
j <- 2
par(mfcol = c(1, 1), pty = "s")
plot(c(-1, 1), c(-1, 1),
     xlab = paste("PC", toString(i)), ylab = paste("PC", toString(j)),
     type = "n",
     main = "Coefficients in the PC-space"
)
arrows(integer(M), integer(M),
       V[, i], V[, j],
       length = .1,
       col = "blue"
)
text(V[, i] * 1.1, V[, j] * 1.1, attributeNames, cex = 0.8)
# Add a unit circle
th <- seq(0, 2.1 * pi, 0.1)
lines(cos(th), sin(th))
```
:::

::: {.col data-latex="{0.05\textwidth}"}
\ 
:::

::: {.col data-latex="{0.40\textwidth}"}
Unfortunately, it is impossible to represent data in a four-dimensional space, so
data are projected onto the bi-dimensional space defined by the first two PCs.
Coefficients can be projected in this space as well, showing the directions
followed by original attributes. It can clearly be seen that the first PC is
eastward oriented, so that it collects very well the eastward attributes.
Probably, the second PC is southward directed so that it collects the southward
attributes (with the plus sign) and the northward *table* attribute (with the
minus sign)
:::
::::::

Data can be projected in all the possible bi-dimensional spaces defined by each
combination of two of the PCs. By projecting them, some insteresting features
can be observed, as shown by the four graphs below.

- Variable *cut* can assume five possible levels, so it is the easiest one to be
visualized. The graph shows that *cut* varies mostly in the direction of the
second PC, so the *cut* of diamonds strongly depends on their quality features
(the better are *color* and *clarity*, the better will be the cut), and is less 
affected by the size (*x*, *y* and *z*) and the *carat* (weight);
- variable *price* is strongly asymmethrical towards low values, and it is
very evident also in the graph where cold blue is predominant with respect to
light blue. *Price* strongly varies with the first PC, so it depends on
quantity factors of diamonds (the bigger and heavier is the diamond, the more 
expensive will be) more than on quality ones;
- variable *carat* is very concentrated between 0 and 1 and widespread beyond 1.
So, in order to have a better visualization, the graph below represent the logarithm
of the *carat*. It can be seen that it is correlated to the dimensional quantities
explained by the first PC (the more expensive and bigger is the diamond, the 
heavier will be) more than quality ones;
- variable *color* has a better representation if data are projected onto the
plan defined by the second and forth PCs. It can be seen that *color* varies
with the forth PC, so it is correlated with *clarity*, *cut* and *table*.

```{r, fig.ncol = 2, out.width = "50%"}
# CUT, PC 1-2
Z <- S$u %*% diag(S$d)
Z <- cbind(Z,diamonds$cut)

i <- 1
j <- 2

ggplot() +
  ggtitle('Cut') +
  geom_point(aes(x = Z[, i], y = Z[, j], color=Z[,11]), size = 1, alpha = 0.5) +
  theme(legend.position = c(0.88, 0.8), legend.title = element_blank()) +
  labs(x = paste('PC ',i), y = paste('PC ',j)) +
  xlim(-5,10) + ylim(-5,7.5)

# PRICE
Z <- S$u %*% diag(S$d)
Z <- cbind(Z,diamonds$price)

i <- 1
j <- 2

ggplot() +
  ggtitle('Price') +
  geom_point(aes(x = Z[, i], y = Z[, j], color=Z[,11]), size = 1, alpha = 0.5) +
  theme(legend.position = c(0.88, 0.8), legend.title = element_blank()) +
  labs(x = paste('PC ',i), y = paste('PC ',j)) +
  xlim(-5,10) + ylim(-5,7.5)

# CARAT
Z <- S$u %*% diag(S$d)
Z <- cbind(Z,diamonds$carat)

i <- 1
j <- 2

ggplot() +
  ggtitle('Carat') +
  geom_point(aes(x = Z[, i], y = Z[, j], color=log(Z[,11])), size = 1, alpha = 0.5) +
  theme(legend.position = c(0.88, 0.8), legend.title = element_blank()) +
  labs(x = paste('PC ',i), y = paste('PC ',j)) +
  xlim(-5,10) + ylim(-5,7.5)

# COLOR
Z <- S$u %*% diag(S$d)
Z <- cbind(Z,diamonds$color)

i <- 2
j <- 4

ggplot() +
  ggtitle('Color') +
  geom_point(aes(x = Z[, i], y = Z[, j], color=Z[,11]), size = 1, alpha = 0.5) +
  theme(legend.position = c(0.88, 0.8), legend.title = element_blank()) +
  labs(x = paste('PC ',i), y = paste('PC ',j)) +
  xlim(-5,10) + ylim(-5,7.5)
```


# 4) Results: what data have shown


